---
title: "Smokes"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(primer.tutorials)
library(tidymodels)
library(broom)
library(marginaleffects)
library(gt)
```

```{r}
smokes |> 
  mutate(age = 5 * (age %/% 5)) |> 
  group_by(age, sex) |> 
  summarize(prop = mean(smoke == "Yes"), .groups = "drop") |> 
  ggplot(aes(x = age, y = prop)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    facet_wrap(~ sex) +
    scale_y_continuous(labels = scales::percent_format())
```

```{r}
#| cache: true
fit_smokes <- logistic_reg(engine = "glm") |> 
  fit(smoke ~ age + sex, data = smokes)
```

```{r}
library(easystats)
check_predictions(extract_fit_engine(fit_smokes))
```

```{r}
tidy(fit_smokes, conf.int = TRUE)
```

```{r}
tidy(fit_smokes, conf.int = TRUE) |> 
  gt() |> 
  fmt_number(columns = c(estimate, conf.low, conf.high), decimals = 3) |> 
  cols_label(
    term = "Term",
    estimate = "Estimate",
    conf.low = "Lower CI",
    conf.high = "Upper CI"
  ) |> 
  tab_header(title = "Logistic Regression: smoke ~ age + sex")
```

```{r}
predictions(extract_fit_engine(fit_smokes),
            newdata = data.frame(age = c(30, 70, 30, 70),
                                 sex = c("Female", "Female", "Male", "Male")))
```

```{r}
avg_comparisons(extract_fit_engine(fit_smokes), variables = "sex")
```

```{r}
plot_predictions(extract_fit_engine(fit_smokes), condition = c("age", "sex"), draw = FALSE) |> 
  ggplot(aes(x = age, y = estimate, color = sex, fill = sex)) +
    geom_line(linewidth = 1) +
    geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, color = NA) +
    scale_y_continuous(labels = scales::percent_format()) +
    labs(
      title = "Age and Sex Predict Ever-Smoker Status",
      subtitle = "Men are consistently more likely than women to have ever smoked, across all ages",
      x = "Age",
      y = "Predicted Probability of Ever-Smoker Status",
      color = "Sex",
      fill = "Sex",
      caption = "Source: NHANES 2009–2012, via primer.tutorials::smokes"
    ) +
    theme_minimal()
```

Smoking behavior in the U.S. has long been shaped by age and sex, with older adults and men historically more likely to have smoked. Using a 1,000-adult sample from NHANES 2009–2012, we ask: what is the difference in the probability of being an ever-smoker between a 30-year-old woman and a 70-year-old man? We model `smoke` as a Bernoulli outcome linked via logistic regression to age and sex, so predicted probabilities come from transforming the linear combination of these covariates through the logit link. The model treats a heavy lifetime smoker and someone who tried 100 cigarettes decades ago identically, even though they're very different from a campaign-targeting perspective. Men have log-odds of being an ever-smoker about 0.54 higher than women, holding nothing else constant — a positive, statistically detectable difference since the interval excludes zero. Each additional year of age is associated with a 0.007 increase in log-odds of being an ever-smoker, but the interval barely crosses zero, so the effect isn't clearly distinguishable from no effect. Adjusting for age, men have log-odds of being an ever-smoker about 0.545 higher than women — nearly unchanged from the sex-only model, confirming age and sex aren't confounding each other. Comparing a 70-year-old man to a 30-year-old woman, we estimate a 13.3-percentage-point higher probability of ever having smoked (95% CI: 7.2% to 19.3%). Our estimate could be wrong for several reasons: the data comes from NHANES 2009–2012, while the question asks about the population today — smoking cohort patterns have shifted enough that the age and sex coefficients may no longer hold. The `smoke` variable also captures lifetime ever-smoker status rather than current smoking, so our model may overstate who a present-day cessation campaign should actually target. If more current data were available, refitting on a recent NHANES cycle would likely produce a smaller estimate, given falling smoking rates nationwide.