---
title: "Smokes"
format: html
execute:
  echo: false
---

Smoking behavior in the U.S. has long been shaped by age and sex, with older adults and men historically more likely to have smoked. Using a 1,000-adult sample from NHANES 2009–2012, we ask: what is the difference in the probability of being an ever-smoker between a 30-year-old woman and a 70-year-old man?

```{r}
#| message: false
library(tidyverse)
library(primer.tutorials)
library(tidymodels)
library(broom)
```

```{r}
smokes |> 
  mutate(age = 5 * (age %/% 5)) |> 
  group_by(age, sex) |> 
  summarize(prop = mean(smoke == "Yes"), .groups = "drop") |> 
  ggplot(aes(x = age, y = prop)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    facet_wrap(~ sex) +
    scale_y_continuous(labels = scales::percent_format())
```

```{r}
#| cache: true
fit_smokes <- logistic_reg(engine = "glm") |> 
  fit(smoke ~ age + sex, data = smokes)
```

```{r}
library(easystats)
check_predictions(extract_fit_engine(fit_smokes))
```


The model treats a heavy lifetime smoker and someone who tried 100 cigarettes decades ago identically, even though they're very different from a campaign-targeting perspective. Men have log-odds of being an ever-smoker about 0.54 higher than women, holding nothing else constant — a positive, statistically detectable difference since the interval excludes zero. 
Each additional year of age is associated with a 0.007 increase in log-odds of being an ever-smoker, but the interval barely crosses zero, so the effect isn't clearly distinguishable from no effect. Adjusting for age, men have log-odds of being an ever-smoker about 0.545 higher than women — nearly unchanged from the sex-only model, confirming age and sex aren't confounding each other.
